Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}

### 2016-04-15 paper/paper.tex
Maximally pruned DBGs are needed for ST functionality

% Graphs
\newcommand{\gindegree}{\ensuremath{\mathsf{in}}}
\newcommand{\goutdegree}{\ensuremath{\mathsf{out}}}
\newcommand{\glabel}{\ensuremath{\mathsf{label}}}
\newcommand{\gkey}{\ensuremath{\mathsf{key}}}
\newcommand{\gvalue}{\ensuremath{\mathsf{value}}}
\newcommand{\gnode}{\ensuremath{\mathsf{node}}}

@@

@@

\begin{definition}[de~Bruijn graph of a graph]
Let $G = (V, E)$ be a labeled graph, and let $\mathcal{S}$ be the \kcollection{k} of maximal path labels in $G$. The \orderk{k} de~Bruijn graph of $\mathcal{S}$ is the \orderk{k} de~Bruijn graph of graph $G$.
\end{definition}

@@

\begin{definition}[Equivalent path graphs]
Let $G = (V, E)$ and $G' = (V', E')$ be two path graphs, and let $k > 0$ be a parameter value. We say that graphs $G$ and $G'$ are \emph{\kequivalent{k}}, if we have $G.\locate(X) = G'.\locate(X)$ for all patterns $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ with $1 \le \abs{X} \le k$.
\end{definition}

\begin{definition}[Pruned de~Bruijn graph]
Let $G = (V, G)$ be a graph, and let $G' = (V', E')$ be an \orderk{k} path graph of $G$. Path graph $G'$ is an \orderk{k} \emph{pruned de~Bruijn graph}, if it is \kequivalent{k} to the \orderk{k} de~Bruijn graph of $G$.
\end{definition}

@@

Now let $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ be a path matching pattern $X$ in graph $G''$. If $\abs{X} = 1$, we can choose any $v_{0} \in V_{d}(v''_{0})$. Now assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$. As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, there must be node $u \in V$ with $G.\glabel(u) = X[j]$ and with an edge $(u, v) \in E$ to a node $v \in G''.\gvalue(v''_{j+1}) = G_{d}.\gvalue(v_{j+1})$. Therefore substring $X[j] \cdot G_{d}.\gkey(v_{j+1})$ exists in $\mathcal{S}$, and the node $v_{j} \in V_{d}$ with $G_{d}.\gkey(v_{j}) = (X[j] \cdot G_{d}.\gkey(v_{j+1}))[0, k-1]$ has $X[j, \abs{X}-1]$ as a prefix of its key. As $G''.\gkey(v''_{j})$ is a prefix of $X[j] \cdot G''.\gkey(v''_{j+1})$, which is a prefix of $X[j] \cdot G_{d}.\gkey(v_{j+1})$, key $G''.\gkey(v''_{j})$ is also a prefix of $G_{d}.\gkey(v_{j})$, and hence $v_{j} \in V_{d}(v''_{j})$. When $j = 0$, we can extend the substring as above to get a path matching the pattern in $G_{d}$, with $v_{0}$ as the start node.

@@

>>>>

% Graphs
\newcommand{\gindegree}{\ensuremath{\mathsf{in}}}
\newcommand{\goutdegree}{\ensuremath{\mathsf{out}}}
\newcommand{\glabel}{\ensuremath{\mathsf{label}}}
\newcommand{\gpred}{\ensuremath{\mathsf{pred}}}
\newcommand{\gkey}{\ensuremath{\mathsf{key}}}
\newcommand{\gvalue}{\ensuremath{\mathsf{value}}}
\newcommand{\gnode}{\ensuremath{\mathsf{node}}}

@@

Given a graph $G = (V, E)$, we may want to reason about the \emph{predecessors} of a node with the given label. Let $v \in V$ be a node and $c \in \Sigma$ be a character. We write $G.\gpred(v, c)$ to denote the set of nodes $u \in V$ such that $G.\glabel(u) = c$ and there is an edge $(u, v) \in E$.

@@

\begin{definition}[de~Bruijn graph of a graph]
Let $G$ be a labeled graph, and let $\mathcal{S}$ be the \kcollection{k} of maximal path labels in $G$. The \orderk{k} de~Bruijn graph of $\mathcal{S}$ is the \orderk{k} de~Bruijn graph of graph $G$.
\end{definition}

@@

\begin{definition}[Equivalent path graphs]
Let $G$ and $G'$ be two path graphs, and let $k > 0$ be a parameter value. We say that graphs $G$ and $G'$ are \emph{\kequivalent{k}}, if we have $G.\locate(X) = G'.\locate(X)$ for all patterns $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ with $1 \le \abs{X} \le k$.
\end{definition}

\begin{definition}[Pruned de~Bruijn graph]
Let $G$ be a graph, and let $G'$ be an \orderk{k} path graph of $G$. Path graph $G'$ is an \orderk{k} \emph{pruned de~Bruijn graph}, if it is \kequivalent{k} to the \orderk{k} de~Bruijn graph of $G$.
\end{definition}

@@

Now let $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ be a path matching pattern $X$ in graph $G''$. If $\abs{X} = 1$, we can choose any $v_{0} \in V_{d}(v''_{0})$. Now assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$. As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, it must be that $\abs{G.\gpred(v, X[j])} > 0$ for a node $v \in G''.\gvalue(v''_{j+1}) = G_{d}.\gvalue(v_{j+1})$. Therefore substring $X[j] \cdot G_{d}.\gkey(v_{j+1})$ exists in $\mathcal{S}$, and the node $v_{j} \in V_{d}$ with $G_{d}.\gkey(v_{j}) = (X[j] \cdot G_{d}.\gkey(v_{j+1}))[0, k-1]$ has $X[j, \abs{X}-1]$ as a prefix of its key. As $G''.\gkey(v''_{j})$ is a prefix of $X[j] \cdot G''.\gkey(v''_{j+1})$, which is a prefix of $X[j] \cdot G_{d}.\gkey(v_{j+1})$, key $G''.\gkey(v''_{j})$ is also a prefix of $G_{d}.\gkey(v_{j})$, and hence $v_{j} \in V_{d}(v''_{j})$. When $j = 0$, we can extend the substring as above to get a path matching the pattern in $G_{d}$, with $v_{0}$ as the start node.

@@

\begin{definition}[Maximally pruned de~Bruijn graph]
Let $G'$ be a a pruned de~Bruijn graph of graph $G$ with key set $\mathcal{K}$. We say that $G'$ is \emph{maximally pruned}, if we cannot prune it using Lemma~\ref{lemma:dbg-prune}, and the keys in $\mathcal{K}$ cannot be made shorter without changing the graph structure.
\end{definition}

\begin{lemma}[Consequences of maximal pruning]\label{lemma:dbg-maximal}
Let $G' = (V', E')$ be a maximally pruned \orderk{k} de~Bruijn graph. Then (a) $\abs{G'.\gpred(v', c)} \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$; and (b) the key $G'.\gkey(v')$ prefix-matches pattern $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ for all nodes $v' \in G'.\find(X)$.
\end{lemma}

\begin{proof}
(a) Let $v' \in V'$ be a node, and let $c \in \Sigma$ be a character. For each node $u' \in G'.\gpred(v', c)$, key $G.\gkey(u')$ prefix-matches string $c \cdot G.\gkey(v')$. If $\abs{G'.\gpred(v', c)} > 1$, the keys must be longer than string $c \cdot G.\gkey(v)$ in order to be distinct. In that case, there cannot be an edge $(u', w')$ to another node $w' \ne v'$, as the keys $G.\gkey(v')$ and $G.\gkey(w')$ would prefix-match. Hence $G'.\gvalue(u')$ must be the union of sets $G.\gpred(v, c)$ over nodes $v \in G'.\gvalue(v')$, and we can merge set $G'.\gpred(v', c)$ into a single node. This contradicts the assumption that graph $G'$ is maximally pruned, so it must be that $\abs{G'.\gpred(v', c)} \le 1$.

(b) If $\abs{X} = 0$, every key prefix-matches the pattern. If $\abs{X} = 1$, set $G'.\find(X)$ is the set of nodes $v' \in V'$ with $G'.\glabel(v') = X[0]$. Now assume that $\abs{X} \ge 2$ and that $G'.\gkey(v')$ prefix-matches $X[1, \abs{X}-1]$ for all $v' \in G'.\find(X[1, \abs{X}-1])$. By (a), there is at most one node $u' \in G'.\gpred(v', X[0])$ for each such node $v'$. As the keys are as short as possible, key $G'.\gkey(u')$ must be a prefix of $X[0] \cdot G'.\gkey(v')$. Because $X[1, \abs{X}-1]$ prefix-matches $G'.\gkey(v')$, pattern $X$ prefix-matches string $X[0] \cdot G'.\gkey(v')$, and hence the pattern must prefix-match key $G'.\gkey(u')$.
\end{proof}
